\documentclass[10pt]{report}

\usepackage[margin=0.3in]{geometry}
\usepackage{multicol}

\usepackage{enumitem}
\setitemize{itemsep=0pt}

\renewcommand{\thesection}{Lec.~\arabic{section}}
\setlength{\parindent}{0pt}


\begin{document}
  \begin{multicols}{2}
    \section{Introduction}
    
    \textbf{Compiler Stages}:
    \begin{enumerate}
      \item \textit{Lexer/scanner}: Read text and produce token stream. FSM-based.  Tools: (F)lex
      \item Parser - Analyze grammatical structure.
      \item Abstract syntax tree \& Symbol Table
      \item Static semantics
      \item Code generation
    \end{enumerate}
  
    \section{Lexical Analysis}
    
    \textbf{Finite State Machine}: Consumes one character on each move and announces a pattern when matched.
    \begin{itemize}
      \item Linear time
      \item Possible States: (Non-)Accepting, Start
    \end{itemize}
  
    \textbf{Angelic Non-determinism}: Try all possible paths. If any ends in a final state, accept.
    
    \textbf{Limitations of Regular Expressions}: Not fully expressive for many regular languages.
    \begin{enumerate}
      \item \textit{Example}: Balanced parentheses
      \item \textit{Intuition}: Regular expressions cannot count.
    \end{enumerate}
  
    \textbf{Representing an NFA as a DFA}: If the NFA has $k$~states, the equivalent DFA will have \textit{at most} $2^{k}$ states.
    
    
    \textbf{Chomsky Hierarchy}: Orders languages from most to least expressive.
    \begin{itemize}
      \item \textbf{Type~0}: No restrictions
      \item \textbf{Type~1}: Context sensitive
      \item \textbf{Type~2}: Context free
      \item \textbf{Type~3}: Right linear, i.e.,~regular
    \end{itemize}
    
    \section{Parsing}
    
    \textbf{LL(*)} -- Left-to-right, leftmost derivation
    \begin{itemize}
      \item Example: ANTLR
      \item Top-down parser
      \item Rely on first and follow sets
      \item Construct right skewed trees
    \end{itemize}
    \textbf{LL1}: Also known as recursive descent parser.
    \textbf{Angelic Recursive Descent Parser}: Example shown in class that similar to an NFA, the angelic parser ``magically'' which production to select at any point in the input.
  
    \textbf{LR(*)}: Left-to-right, rightmost derivation
    \begin{itemize}
      \item Examples: Bison, Yacc, CUP 
      \item Bottom-Up parser: Build leaves then parents
      \item Construct left skewed trees
    \end{itemize}
  
    Two parser operators:
    \begin{enumerate}
      \item \textit{Peak}: Check the next character and report an error if it is invalid.
      \item \textit{Match}: Consume next token if it is the same as the expected value or throw an error.
    \end{enumerate}
  
    \textbf{First Set}: Set of terminals that can appear at the beginning of a specific non-terminal. \textbf{Follow Set}: Set of possible terminals can that appear after a specific non-terminal.
    
    \textbf{Example Grammar with Precedence}:
    
    \texttt{% 
    expr $\rightarrow$ expr '+' term\\
    \hbox{\hspace{30pt}}| expr '-' term\\
    \hbox{\hspace{30pt}}| term\\
    \\
    term $\rightarrow$ term '*' factor\\
    \hbox{\hspace{30pt}}| term '/' factor\\
    \hbox{\hspace{30pt}}| factor\\
    \\
    factor $\rightarrow$ '-' factor\\
    \hbox{\hspace{39pt}}| prim\\
    \\
    prim $\rightarrow$ NUMBER}
    
    \subsection{Bottom-Up Parser Operations}
    
    \textbf{Shift}: Move over a token
    
    \textbf{Reduce}: At end of rule, reduce to non-terminal then do a GoTo.
    
    \textbf{GoTo}: Move over non-terminal. \textit{Can never have a conflict}
    
    \textbf{Dotted Item}: Represents how far along we are in the production. \textbf{Kernel}: Set of productions yielded by applying the transition operation.
    
    \subsection{Resolving Conflicts w/ Lookahead}
    \textbf{Shift-Reduce}: Character after shift not in reduce lookahead.
    
    \textbf{Reduce/Reduce}: Intersection of all pairs reduce lookaheads empty.
    
    \section*{Languages}
    
    \noindent
    \textbf{Regular Language}: Any language that can be represented via a DFA/NFA or that can be expressed via a regular expression.
    \begin{itemize}
      \item All finite languages are inherently regular.
    \end{itemize}
    
  \end{multicols}
\end{document}